\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex
Tweaks

Outside of the context of simulation, ARG-unary nodes are
only like to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-Cov-2; see Discussion).
\emph{Locally} unary nodes, those which have one child
in one or more local trees, are a common and important feature
of many different types of ARG, and a recurring topic
in this article. Without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which nodes can be unary in the local trees
of an ARG. Furthermore, ARGs inferred by current methods contain
locally unary nodes from several of them (see section XXX).

@@

\section{Cell lineages and ARGs}
\label{cell-lineages-and-args}
As we have argued, an important benefit of the gARG representation is that it
can concatenate multiple events by encoding the outcome of their effects
on a single node. In the case of recombination events, this manifests itself as
a single node with three or more parents (e.g.\ Fig.~\ref{fig-simplification}C, node \noderef{e}).
Here we make the argument that the same principle underlies ARGs containing
coalescence points that involve three or more children. These ``polytomies'' are generated not only
by Wright-Fisher dynamics such as those in Fig.~\ref{fig-simplification}C (see node \noderef{k}), but also
by life histories that involve ``burst'' reproduction events, which are often modelled by a lambda
coalecent process [AND OTHERS]\citep{wakely2008coalescent}.

@@

At the base of this concept is a subtle and important point about which genomes the nodes in a gARG represent. 
In principle, we can interpret nodes as representing genomes
at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
In reality, however, recombination and replication of DNA occur at a cellular level,
and correspond to the processes of meiosis and mitosis respectively. A comprehensive gARG would therefore
trace the process of inheritance not between diploid individuals, but between each cell during the
life cycle of an organism. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
schematic of cellular inheritance in a diploid population over four partially overlapping generations.
In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
four sampled chromosomes. The potential ARG nodes are shown as green chromosomal regions, and the ``events''
underlying the ARG are shown as red and blue squares.

The unusual feature of this particular example is that three ARG lineages coalesce in one individual, $D_{10}$
(who has three children). We know that the underlying coalescence process must consist of two successive bifurcations
as the only known method of reproducing DNA is by (semi-conservative) duplication. However, if we take each node in the
ARG to represents the generalised genome of an individual, rather than the specific genome of a cell within an individual,
we can only represent this as a instantaneous trifurcation or polytomy.

@@

As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
by a polytomy in the local trees.

>>>>

Outside of the context of simulation, ARG-unary nodes are
only like to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-Cov-2; see Discussion).
\emph{Locally} unary nodes, those which have one child
in one or more local trees, are a common and important feature
of many different types of ARG, and a recurring topic
in this article. Without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Locally unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain locally unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which ARG nodes can be locally unary. Furthermore, 
several current ARG inference methods infer locally unary nodes (see section XXX).

@@

\section{Cell lineages and ARGs}
\label{cell-lineages-and-args}
As we have argued, an important benefit of the gARG representation is that it
can concatenate multiple events by encoding the outcome of their effects
on a single node. In the case of recombination events, this manifests itself as
a single node with three or more parents (e.g.\ Fig.~\ref{fig-simplification}C, node \noderef{e}).
Here we make the argument that the same principle underlies ARGs containing
coalescence points that involve three or more children. These ``polytomies'' are generated not only
by Wright-Fisher dynamics such as those in Fig.~\ref{fig-simplification}C (see node \noderef{k}), but also
by life histories that involve ``burst'' reproduction events, which are often modelled by a lambda
coalescent process [AND OTHERS]\citep{wakely2008coalescent}.

@@

At the base of this concept is a subtle and important point about which genomes the nodes in a gARG represent. 
In principle, we can interpret nodes as representing genomes
at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
In reality, however, recombination and replication of DNA occur at a cellular level,
and correspond to the processes of meiosis and mitosis respectively. A comprehensive gARG would therefore
trace the process of inheritance not between diploid individuals, but between each cell during the
life cycle of an organism. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
schematic of cellular inheritance in a diploid population over four partially overlapping generations.
In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
four sampled chromosomes. The ``events'' underlying the ARG are shown as red and blue squares,
and the potential ARG nodes are shown as green chromosomal regions. Note that a number of these nodes are
``pass-through'' nodes with one parent and one child, and in practice these would normally be omitted.

The unusual feature of this particular example is that three ARG lineages coalesce in one historical individual,
$D_10$, whose three children all happen to be ancestors of the sampled chromosomes.
We can see that the process that generates these three child lineages actually consists of two successive bifurcations:
necessitated by the fact that the only known method of reproducing DNA is by (semi-conservative) duplication.
However, if we take each node in the ARG to represents the generalised genome of an individual (rather than the specific
genome of a cell within an individual) we end up representing this as a instantaneous trifucation or polytomy.

@@

% As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
% coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
% other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
% a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
% eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
% by a polytomy in the local trees.
